Lesson 5
========================================================

### Multivariate Data
Notes:

***

### Moira Perceived Audience Size Colored by Age
Notes:

***

### Third Qualitative Variable
Notes:

```{r Third Qualitative Variable}
library(ggplot2)
setwd("/home/jonas/workspace/udacity/lessons/05 - Exploratory Data Analysis/")
pf <- read.csv("pseudo_facebook.tsv", sep="\t")
ggplot(aes(x = gender, y = age),
       data = subset(pf, !is.na(gender))) + 
  geom_histogram()

library(dplyr)
by_age_and_gender <- group_by(pf, age, gender)
pf.fc_by_age_gender <- summarize(by_age_and_gender,
                                 mean_friend_count = mean(friend_count),
                                 median_friend_count = median(friend_count),
                                 n = n())
head(pf.fc_by_age_gender)
# Alternative syntax:
pf.fc_by_age_gender <- pf %>%
  group_by(age, gender) %>%
  summarize(mean_friend_count = mean(friend_count),
            median_friend_count = median(friend_count),
            n = n()) %>%
  ungroup() %>%
  arrange(age)

ggplot(data = subset(pf.fc_by_age_gender, !is.na(gender)),
       aes(x = age, y = median_friend_count)) +
  geom_line(aes(color = gender))

```

***

### Plotting Conditional Summaries
Notes:

```{r Plotting Conditional Summaries}

```

***

### Thinking in Ratios
Notes:

***

### Wide and Long Format
Notes:

***

### Reshaping Data
Notes:

```{r}
install.packages('reshape2')
library(reshape2)
pf.fc_by_age_gender.wide <- dcast(pf.fc_by_age_gender,
                                  age ~ gender,
                                  value.var = "median_friend_count")  # syntax is (keep1 + keep2 + ... ~ distinguish)
# inverse of dcast() is melt()

head(pf.fc_by_age_gender.wide)

# alternativ solution using tidyr and dplyr
library(tidyr)
#pf.fc_by_age_gender.wide <-
#  subset(pf.fc_by_age_gender[c('age', 'gender', 'median_friend_count')],
#         !is.na(gender)) %>%
#  spread(gender, median_friend_count) %>%
#  mutate(ratio = male / female)

```


***

### Ratio Plot
Notes:

```{r Ratio Plot}
ggplot(data = pf.fc_by_age_gender.wide, 
       aes(x = age, y = female / male)) +
  geom_line() + 
  geom_hline(yintercept = 1, linetype = 2, alpha = .3)


```

***

### Third Quantitative Variable
Notes:

```{r Third Quantitative Variable}
pf$year_joined <- floor(2014 - pf$tenure / 365)
```

***

### Cut a Variable
Notes:

```{r Cut a Variable}
summary(pf$year_joined)
table(pf$year_joined)
pf$year_joined.bucket <- cut(pf$year_joined, breaks = c(2004, 2009, 2011, 2012, 2014))
head(pf$year_joined.bucket)
```

***

### Plotting it All Together
Notes:

```{r Plotting it All Together}
table(pf$year_joined.bucket, useNA = "ifany")
subset(pf, !is.na(year_joined.bucket)) %>% 
  ggplot(aes(x = age, y = friend_count)) +
  geom_line(aes(color = year_joined.bucket), stat = "summary", fun.y = median)
```

***

### Plot the Grand Mean
Notes:

```{r Plot the Grand Mean}

subset(pf, !is.na(year_joined.bucket)) %>% 
  ggplot(aes(x = age, y = friend_count)) +
  geom_line(aes(color = year_joined.bucket), stat = "summary", fun.y = mean) + 
  geom_line(stat = "summary", fun.y = mean, linetype = 2)
```

***

### Friending Rate
Notes:

```{r Friending Rate}
with(subset(pf, tenure > 0), summary(friend_count / tenure))

```

***

### Friendships Initiated
Notes:

What is the median friend rate?

What is the maximum friend rate?

```{r Friendships Initiated}
ggplot(data = subset(pf, pf$tenure > 0),
       aes(x = tenure, y = friendships_initiated / tenure)) +
  geom_line(aes(color = year_joined.bucket), stat = "summary", fun.y = mean)
```

***

### Bias-Variance Tradeoff Revisited
Notes:

```{r Bias-Variance Tradeoff Revisited}

ggplot(aes(x = tenure, y = friendships_initiated / tenure),
       data = subset(pf, tenure >= 1)) +
  geom_line(aes(color = year_joined.bucket),
            stat = 'summary',
            fun.y = mean)

ggplot(aes(x = 7 * round(tenure / 7), y = friendships_initiated / tenure),
       data = subset(pf, tenure > 0)) +
  geom_line(aes(color = year_joined.bucket),
            stat = "summary",
            fun.y = mean)

ggplot(aes(x = 30 * round(tenure / 30), y = friendships_initiated / tenure),
       data = subset(pf, tenure > 0)) +
  geom_line(aes(color = year_joined.bucket),
            stat = "summary",
            fun.y = mean)

ggplot(aes(x = 90 * round(tenure / 90), y = friendships_initiated / tenure),
       data = subset(pf, tenure > 0)) +
  geom_line(aes(color = year_joined.bucket),
            stat = "summary",
            fun.y = mean)


ggplot(aes(x = tenure, y = friendships_initiated / tenure),
       data = subset(pf, tenure >= 1)) +
  geom_smooth(aes(color = year_joined.bucket))
```

***

### Sean's NFL Fan Sentiment Study
Notes:

***

### Introducing the Yogurt Data Set
Notes:

***

### Histograms Revisited
Notes:

```{r Histograms Revisited}
yo <- read.csv("yogurt.csv")
str(yo)
head(yo)
yo$id <- factor(yo$id)
str(yo)

ggplot(data = yo, aes(price)) +
  geom_histogram(binwidth = 1)
```

***

### Number of Purchases
Notes:

```{r Number of Purchases}

yo <- transform(yo, all.purchases = strawberry + blueberry + pina.colada + plain + mixed.berry)
colnames(yo)
head(yo)
```

***

### Prices over Time
Notes:

```{r Prices over Time}
ggplot(data = yo, aes(all.purchases)) +
  geom_histogram(color= "blue", binwidth = 1)

ggplot(data = yo, aes(x = time, y = price)) +
  geom_jitter(alpha = .25, shape = 21, fill = "orange")
```

***

### Sampling Observations
Notes:

***

### Looking at Samples of Households

```{r Looking at Sample of Households}
set.seed(4242)
sample.ids <- sample(levels(yo$id), 16)
sample.ids

ggplot(data = subset(yo, id %in% sample.ids),
       aes(x = time, y = price)) +
  facet_wrap( ~ id) +
  geom_line() + 
  geom_point(aes(size = all.purchases), pch = 1)
```

***

### The Limits of Cross Sectional Data
Notes:

***

### Many Variables
Notes:

***

### Scatterplot Matrix
Notes:
```{r}
#install.packages("GGally")
library(GGally)
theme_set(theme_minimal(20))

set.seed(1836)
pf_subset <- pf[, c(2:15)]
names(pf_subset)
ggpairs(pf_subset[sample.int(nrow(pf_subset), 1000), ])  # takes a long time
```
***

### Even More Variables
Notes:

***

### Heat Maps
Notes:

```{r}
nci <- read.table("nci.tsv")
colnames(nci) <- c(1:64)
```

```{r}
nci.long.samp <- melt(as.matrix(nci[1:200,]))
names(nci.long.samp) <- c("gene", "case", "value")
head(nci.long.samp)

ggplot(aes(y = gene, x = case, fill = value),
  data = nci.long.samp) +
  geom_tile() +
  scale_fill_gradientn(colours = colorRampPalette(c("blue", "red"))(100))
```


***

### Analyzing Three of More Variables
Reflection:

***

Click **KnitHTML** to see all of your hard work and to have an html
page of this lesson, your answers, and your notes!

